---
title: "fars_details_markdown"
author: "attackgnome"
date: "May 1, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## fars

This package contains functions to read, summarize, and plot the data available from the NHTSA Fatality Analysis reporting system. For more information on the underlying data, see <https://www.nhtsa.gov/research-data/fatality-analysis-reporting-system-fars>.

##fars_read

To read in the raw data pulled from the FARS website and put it into a data frame tbl, use the fars_read function with the filename entered as the argument. For example, If you want to read in the data from 2014, input the following:


```{r fars_read, echo=FALSE}
fars_read <- function(filename) {
  if(!file.exists(filename))
    stop("file '", filename, "' does not exist")
  data <- suppressMessages({
    readr::read_csv(filename, progress = FALSE)
  })
  dplyr::tbl_df(data)
}
```
```{r fars_read_example}
fars_read("accident_2014.csv.bz2")
```

##make_filename

if you want to create a file for a specific year using the same naming convention as the NHTSA, input the desired year as an argument into the make_filename function. For example if I want to know the file name the NHTSA uses for the data from 2014, you can input the following into the R console:

```{r make_filename, echo=FALSE}
make_filename <- function(year) {
  year <- as.integer(year)
  sprintf("accident_%d.csv.bz2", year)
}
```
```{r make_filename_example}
make_filename(2014)
```

##fars_read_years

Sometimes though its not enough to read in the data for any one single year. If you want to read in a series of years and place it into a data frame tbl, you can use the fars_read_years function with the series of years desired as the argument.
For example, if I wanted to read in the data for 2013 and 2015:

```{r fars_read_years, echo=FALSE}
library(dplyr)

make_filename <- function(year) {
  year <- as.integer(year)
  sprintf("accident_%d.csv.bz2", year)
}


fars_read_years <- function(years) {
  lapply(years, function(year) {
    file <- make_filename(year)
    tryCatch({
      dat <- fars_read(file)
      dplyr::mutate(dat, year = year) %>%
        dplyr::select(MONTH, year)
    }, error = function(e) {
      warning("invalid year: ", year)
      return(NULL)
    })
  })
}
```
```{r fars_read_years_example}
fars_read_years(c(2013,2015))
```

##fars_summarize_years

If you want to summarize the data from a single year or series of years, simply input the year (or series of years as a vector) as an argument into the fars_summarize_years function. For example, if I want to summarize the data for 2013 through 2015, you can do so with the following input:

```{r fars_summarize_years, echo=FALSE}
library(magrittr)
fars_summarize_years <- function(years) {
  dat_list <- fars_read_years(years)
  dplyr::bind_rows(dat_list) %>%
    dplyr::group_by(year, MONTH) %>%
    dplyr::summarize(n = n()) %>%
    tidyr::spread(year, n)
}
```
```{r fars_summarize_years_example}
fars_summarize_years(2013:2015)
```

## fars_map_state

You can also plot the accidents for a single or series of years on a state by state basis. For example, if you want to plot accidents in Alabama (state.num = 1) in 2014:

```{r fars_map_state, echo=FALSE}
library(maps)
library(graphics)
fars_map_state <- function(state.num, year) {
  filename <- make_filename(year)
  data <- fars_read(filename)
  state.num <- as.integer(state.num)

  if(!(state.num %in% unique(data$STATE)))
    stop("invalid STATE number: ", state.num)
  data.sub <- dplyr::filter(data, STATE == state.num)
  if(nrow(data.sub) == 0L) {
    message("no accidents to plot")
    return(invisible(NULL))
  }
  is.na(data.sub$LONGITUD) <- data.sub$LONGITUD > 900
  is.na(data.sub$LATITUDE) <- data.sub$LATITUDE > 90
  with(data.sub, {
    maps::map("state", ylim = range(LATITUDE, na.rm = TRUE),
              xlim = range(LONGITUD, na.rm = TRUE))
    graphics::points(LONGITUD, LATITUDE, pch = 46)
  })
}
```
```{r fars_map_state_example}
fars_map_state(1, 2014)
```

